---
title: "BERD tutorial analysis - WGCNA"
author: "Ellen C. Francis"
date: "Updated June 10 2025"
output:
  html_document:
    code_folding: hide
  pdf_document: default
funding: ECF NICHD R00108272
editor_options:
  markdown:
    wrap: 72
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r setup2, include=FALSE}
#set to your working directory here if you want
knitr::opts_knit$set(root.dir = "C:/Users/ef447/OneDrive - Rutgers University/DOCUMENTS/TALKS/BERD/Tutorial")
```

*Funding: NICHD R00108272*

\# \# Code chunk 1 \#

```{r, Display the current working directory}
setwd("C:/Users/ef447/OneDrive - Rutgers University/DOCUMENTS/TALKS/BERD/Tutorial")
getwd();
# If necessary, change the path below to the directory where the data files are stored. 
# "." means current directory. On Windows use a forward slash / instead of the usual \.
workingDir = ".";
setwd(workingDir); 
# Load the WGCNA package
library(WGCNA)

# Load data wrangling package
library(tidyverse)
library(tibble)

#load packages for association testing
library(CorLevelPlot)
#devtools::install_github("kevinblighe/CorLevelPlot")
library(rstatix)
library(reshape2)
#install.packages("ggstatsplot")
library(ggstatsplot)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE);
# Allow multi-threading within WGCNA. At present this call is necessary.
# Any error here may be ignored but you may want to update WGCNA if you see one.
# Caution: skip this line if you run RStudio or other third-party R environments.
# See note above.
enableWGCNAThreads()
########load metabolomics data########
# (technical filtering has been applied to this data)
bd1_resultsd4 <- read.csv("bd1_resultsd4.csv")
bd1_resultsd4 %>% select(-X) -> bd1_resultsd4 #remove random x variable
as_tibble(bd1_resultsd4)
```

#pivot and transform data so that it conforms to WGCNA rules
#Samples in Rows and Metabolites in Columns

```{r, transform data and inspect}
datExpr <- bd1_resultsd4 %>% pivot_longer(!RT_mass, names_to="sample_name") %>%
  pivot_wider(names_from="RT_mass", values_from="value") %>% 
  column_to_rownames(var = "sample_name") # this is the full metabolomics dataset with all the samples 
as_tibble(datExpr)

#names(df1) # to get the column names
dim(datExpr) #to get the dimensions
```

**QC.**

**Identify potential outliers.**

**Table of WGNCA function to detect outlier "genes"/metabolites and
outlier samples.**

All metabolites and samples ok so we move to normalization

We conduct QC to identify outliers. We exclude three samples because
they look like outliers

```{r qc - find outliers}
gsg <- goodSamplesGenes(datExpr)
summary(gsg)
gsg$allOK
#look to see how many are outliers. All metabolites and samples ok.
table(gsg$goodGenes)
table(gsg$goodSamples)

#detect outliers using hierarchical clustering 

htree <- hclust(dist(datExpr), method ="average")
plot(htree) # potentially some outliers 


# detect outliers using pca
pca <- prcomp(datExpr)
pca.dat <- pca$x
pca.var <- pca$sdev^2
pca.var.percent <- round(pca.var/sum(pca.var*100), digits =2)

pca.dat <- as.data.frame(pca.dat)

ggplot(pca.dat, aes(PC1, PC2)) +
  geom_point() +
  geom_text(label = rownames(pca.dat)) +
  labs(x = paste0('PC1: ', pca.var.percent[1], ' %'),
       y = paste0('PC2: ', pca.var.percent[2], ' %'))

samples.excluded <- c("PZDa_S_603", "PZDa_S_1390", "PZDa_S_2118")  #Possibly exclude

```

We could exclude three samples because they look like outliers
"PZDa_S_603", "PZDa_S_1390", "PZDa_S_2118". Exclusion of these does not
change the results

\# \# Code chunk 2 \#

```{r call_hclust, fig.width=9, fig.height=5,  echo=TRUE}
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
If R² is consistently below 0.8: Your network may not be scale-free, suggesting you might need to try a different power range or consider whether WGCNA is appropriate for this dataset.

Low Mean Connectivity: If mean connectivity is very low, your modules might be poorly defined, indicating a need for a lower power.

*CONNECTIVITY
- High Mean Connectivity: Suggests that many nodes are interconnected, forming dense modules that are likely to reflect real biological networks.
- Low Mean Connectivity: Indicates that most nodes are weakly connected, potentially forming sparse or poorly defined modules, which can make it difficult to identify biologically meaningful clusters.

*POWER
- Higher Power: Makes the adjacency matrix more binary (sharper separation of strong vs. weak connections), reducing overall connectivity.
- Lower Power: Produces a more interconnected network, increasing overall connectivity.

\# \# Code chunk 3 \#

```{r, soft power}
softPower = 6;
adjacency = adjacency(datExpr, power = softPower);
```

\# \# Code chunk 4 \#

```{r, Turn adjacency into topological overlap}
TOM = TOMsimilarity(adjacency);
dissTOM = 1-TOM
```

\# \# Code chunk 5 \#

```{r Call the hierarch func, fig.width=12, fig.height=9, echo=TRUE}
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Metabolite clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04);

```

\# \# Code chunk 6 \#

```{r, specify module size}
#We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                deepSplit = 2, pamRespectsDendro = FALSE,
                minClusterSize = minModuleSize);
table(dynamicMods)
```

\# \# Code chunk 7 \#

```{r, Convert numeric lables into colors, fig.width=8, fig.height=6, echo=TRUE}
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Metabolite dendrogram and module colors")
```

\# \# Code chunk 8 \#

```{r, Calculate eigengenes and change cut threshold, fig.width=7, fig.height=6, echo=TRUE}
MEList = moduleEigengenes(datExpr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
plot(METree, main = "Clustering of module eigenmetabolites",
     xlab = "", sub = "")

## change this number to change the cutting threshold
MEDissThres = 0.4
MEDissThresHigh = 0.6

# Plot the dendrogram
plot(METree, main = "Clustering of Module Eigengenes",
     xlab = "", sub = "")

# Add the first cut line (low threshold)
abline(h = MEDissThres, col = "red", lty = 2, lwd = 2)  # Dashed red line

# Add the second cut line (high threshold)
abline(h = MEDissThresHigh, col = "blue", lty = 1, lwd = 2)  # Solid blue line

legend("topright", legend = c("Low Threshold (9 MEs)", "High Threshold (6 MEs)"),
       col = c("red", "blue"), lty = c(2, 1), lwd = 2)


```

-   This height threshold merges modules that are clustered together
    with a distance below this line.

    \# \# Code chunk 9 \#

```{r, merge modules}
# Call an automatic merging function
merge = mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs;
```

\# \# Code chunk 10 \#

```{r, dendogram plots, fig.width=12, fig.height=9, echo=TRUE}
#pdf(file = "Plots/geneDendro-3.pdf", wi = 9, he = 6)
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
table(mergedColors)
table(dynamicColors)

#dev.off()
```

\# \# Code chunk 11 \#

```{r, names modules colors}
# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;
# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors, geneTree, file = "Tutorial_Network.RData")

glimpse(MEs)
```

Determine which metabolites are hubs by determine which metabolites have
extremely high kME values. ***HUB METABOLIES***

\# \# Code chunk 12 \#

```{r determine hubs eigen modules}
dim(datExpr)  # Should be (samples x metabolites)
dim(MEs)      # Should be (samples x modules)
glimpse(rownames(datExpr))
glimpse(rownames(MEs))
# Check if the rownames match
all(rownames(MEs) == rownames(datExpr))

# Calculate module membership (kME) for each metabolite
geneModuleMembership = signedKME(datExpr, MEs)

# Use the module names from MEs for consistent labeling
colnames(geneModuleMembership) = paste0("k_", names(MEs))

# Calculate p-values for each kME
MMPvalue = corPvalueStudent(as.matrix(geneModuleMembership), nrow(datExpr))
colnames(MMPvalue) = paste0("pval_", names(MEs))

# Extract gene (metabolite) names (use column names of datExpr)
Gene = colnames(datExpr)

# Create the full kME table
kMEtable = data.frame(Metabolite = Gene, 
                      Module = moduleColors,  # Correct module assignments
                      geneModuleMembership, 
                      MMPvalue)

# Save the full kME and p-value table to CSV
write.csv(kMEtable, "kMEtable.csv", row.names = FALSE)

# Preview the first few rows to confirm - this is a matrux of metabolites and their modules
glimpse(kMEtable)
```

\# \# Code chunk 13 \#

\`

```{r determine hubs in networks}
# Initialize list to store top hub genes per module
topGenesKME = list()

# Extract module names from the kME matrix
module_names = colnames(geneModuleMembership)

# Loop through each module to identify top hub genes
for (module in module_names) {
  
  # Rank genes within the module (higher kME is better for hubs)
  kMErank = rank(-geneModuleMembership[, module])
  
  # Extract the top 10 hub genes
  topGenes = Gene[kMErank <= 10]
  
  # Store in the list
  module_name_cleaned = sub("kME_", "", module)
  topGenesKME[[module_name_cleaned]] = topGenes
}

# Convert the list to a data frame for easier viewing
topGenesKME_df = as.data.frame(topGenesKME)

# Save the top hub genes to a CSV file
write.csv(topGenesKME_df, "TopGenesKME.csv", row.names = FALSE)

# Preview the top hub genes
print(topGenesKME_df)

```

\# \# Code chunk 14 \#

```{r, peak intensity module}
# Loop through each module
for (module in unique(mergedColors)) {
  
  # Extract the ME for the current module
  ME_values = MEs[, paste0("ME", module)]
  
  # Order the samples by ME values
  ordered_indices = order(ME_values)
  ordered_ME_values = ME_values[ordered_indices]
  
  # Create the bar plot
  bp = barplot(ordered_ME_values, 
               col = module, 
               main = paste("Module:", module), 
               cex.main = 0.8, 
               ylab = "Eigengene Expression", 
               xlab = "Sample", 
               border = NA, 
               las = 2)
  
  # Calculate the 25th, 50th, and 75th percentiles
  n_samples = length(ordered_ME_values)
  quantiles = quantile(1:n_samples, probs = c(0.25, 0.5, 0.75))
  
  # Add vertical lines at the quantiles
  abline(v = bp[quantiles], col = "black", lty = 2, lwd = 2)
  
  # Add percentile labels
  text(x = bp[quantiles], y = max(ordered_ME_values) * 0.95, 
       labels = c("25%", "50%", "75%"), 
       pos = 3, cex = 0.7, col = "black")
}
```

-   Code chunk 15

-   correlations with clinical data. This will not work because of data
    access issues

```{r correlation with traits.outcomes, echo=FALSE, results='hide'}
# read in imputed outcomes and traits data to use for clinical insight
read.csv("trts.outcms.impscl.csv") -> trts.outcms.impscl
#row indictor is removed
trts.outcms.impscl %>% select(-X) -> trts.outcms.impscl

#remove imputed prefix indicators for simplicity
colnames(trts.outcms.impscl)<-gsub("imputed_cart.","",colnames(trts.outcms.impscl))
colnames(trts.outcms.impscl)<-gsub("imputed_pmm.","",colnames(trts.outcms.impscl))

#read in sample manifest
sbd1_factorsd2 <- read.csv("sbd1_factorsd2.csv")

#1065 samples to be selected from the ipv1 samples -> this works bc the covariates don't vary over pregnancy 
#so we can link based on PID and not sample
trts.outcms.impscl %>% filter(pid %in% sbd1_factorsd2$PID) -> trts.outcms.impscl

trts.outcms.impscl.1 <- sbd1_factorsd2 %>% select(sample_name, PID) %>% 
  mutate(pid=PID) %>% 
  right_join(trts.outcms.impscl, by="pid") 

trts.outcms.impscl.1 <- trts.outcms.impscl.1 %>% select(-PID, -pid)

arrange(trts.outcms.impscl.1, desc(sample_name))
```

\*sample of the clinical characteristics available.

**Heat map of the correlation between metabolites and covariates**

We check these associations to assess whether we should be dropping any
modules or focusing in on certain modules.

```{r add modules to clinical data}
#check if row names and IDs are the same
identical(rownames(MEs), trts.outcms.impscl.1$sample_name)
#more data chekcs
nSamples <- nrow(datExpr)
nMets <- ncol(datExpr)

#move sample names to row names and drop
trts.outcms.impscl.1 <- column_to_rownames(trts.outcms.impscl.1, var="sample_name")

#pearson correlation
module.covars.corr <- cor(MEs, trts.outcms.impscl.1, use = "p")
module.covars.corr.pvals <- corPvalueStudent(module.covars.corr, nSamples)
glimpse(module.covars.corr.pvals)
```

```{r heatmap covars and metabolites,  out.width = "80%", fig.width=15, fig.height=8}
library(CorLevelPlot)
covars.heatmap.data <- merge(MEs, trts.outcms.impscl.1, by = "row.names")
glimpse(covars.heatmap.data) # i use this to see what I want on the X axis and Y axis. We have 9 MEs so i'll but them on X
dim(covars.heatmap.data)
#move ids back to row name
covars.heatmap.data <-covars.heatmap.data %>%
  column_to_rownames(var = "Row.names")

CorLevelPlot(covars.heatmap.data, 
             x =names(covars.heatmap.data)[1:9],
             y= names(covars.heatmap.data)[c(10:17,20)], #if you just want to select certain variables
             col = c("blue1", "skyblue", "white", "pink", "Red"),
            cexLabX = .75,
            rotLabX = 45,
            colLabX = "grey20",
            fontLabX = 1,
            cexLabY = .75,
            fontLabY = 1,
        cexCorval = .75,
        fontCorval = .5,
        cexLabColKey = .75)
```

\`\`

```{r compare expression levels by group}

# Extract only the relevant ME columns and the grouping variable
ME_colors <- c("MEpink", "MEblack", "MEbrown", "MEgreen", "MEyellow", "MEpurple")
ME.exprs <- select(covars.heatmap.data, all_of(ME_colors), rcluster_5)

# Define color palettes for each module
color_palettes <- c(
  "MEpink" = "PiYG",
  "MEblack" = "Greys",
  "MEred" = "RdGy",
  "MEgreen" = "Greens",
  "MEyellow" = "YlOrBr",
  "MEbrown" = "PuOr",
  "MEblue" = "Blues",
  "MEturquoise" = "YlGnBu",
  "MEpurple" = "Purples"
)

# Generate plots using a loop
plot_list <- list()

for (ME in ME_colors) {
  
  plt <- ggbetweenstats(
    data = ME.exprs,
    x = rcluster_5,
    y = !!sym(ME),
    type = "nonparametric",
    plot.type = "box",
    pairwise.comparisons = TRUE,
    pairwise.display = "s",
    p.adjust.method = "fdr",
    var.equal = FALSE,
    centrality.plotting = FALSE,
    bf.message = FALSE,
    package = "RColorBrewer",
    palette = color_palettes[ME]
  )
  
  # Store the plot in the list
  plot_list[[ME]] <- plt
  
  # Print each plot (optional)
  print(plt)
}

# Access individual plots if needed
plot_list$MEyellow
```


